---
title: "Getting started with ellmercodex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ellmercodex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`ellmercodex` connects `ellmer` text chats to observed, subscription-backed
Codex authentication and response behavior. Its stable core is the public
`ellmer` 0.4.2 `Chat` object for interactive,
single-conversation work. The direct OAuth and response transport are
undocumented compatibility surfaces, may change without notice, and are not
endorsed or supported by OpenAI.

Package loading, examples, vignettes, and automated tests do not authenticate,
open a browser, inspect stored credentials, or make network requests.

## Check the local integration

```{r}
library(ellmercodex)

codex_available()
```

`codex_available()` checks installed dependencies, static configuration, and
the supported `ellmer` compatibility seam. Its default does not inspect a
credential store or contact the network, and `TRUE` does not guarantee that a
remote request will succeed. After authentication, `codex_models()` queries
the observed account-specific catalog and returns the models and reasoning
efforts the account currently advertises. The package supplies its own verified
compatibility value for the catalog query; set `ELLMERCODEX_CLIENT_VERSION`
only when a different explicit value is needed:

```{r, eval = FALSE}
models <- codex_models()
models[c("id", "display_name", "default_reasoning_effort",
         "supported_reasoning_efforts")]
```

## Authenticate explicitly

Authentication happens only after a direct call to `codex_login()`. The
function opens a browser, validates a loopback callback using PKCE and OAuth
state, and stores the resulting credential in a package-specific OS-keyring
entry by default.

```{r, eval = FALSE}
auth <- codex_login()
codex_account(auth)
```

`codex_account()` omits tokens and redacts the account identifier. The package
does not read the Codex CLI credential file or another application's stored
session.

Set `persist = FALSE` to keep the credential only in the current R process:

```{r, eval = FALSE}
codex_login(persist = FALSE)
chat <- chat_codex(model = Sys.getenv("ELLMERCODEX_MODEL"))
```

The process-local session is cleared by `codex_logout()` and is never written
to disk or the OS keyring.

The OS keyring is the default backend. A file fallback is available only when
both of these environment variables are configured explicitly:

```{r, eval = FALSE}
Sys.setenv(
  ELLMERCODEX_CREDENTIAL_BACKEND = "file",
  ELLMERCODEX_CREDENTIAL_PASSPHRASE = "use-a-secret-from-your-password-manager"
)
```

The fallback is encrypted and stored below
`tools::R_user_dir("ellmercodex", "data")`. No plaintext fallback is provided.

## Create a text chat

By default, `chat_codex()` selects a usable model from the authenticated
account catalog. Select a model explicitly when you need a stable choice, and
choose reasoning effort from that model's advertised values:

```{r, eval = FALSE}
chat <- chat_codex(
  system_prompt = "Be concise.",
  model = models$id[[1]],
  effort = models$supported_reasoning_efforts[[1]][[1]]
)

chat$chat("Remember that the codeword is amber. Reply: acknowledged")
chat$chat("What was the codeword?")
```

The returned object is an `ellmer` 0.4.2 `Chat`. The complete public Chat
surface remains available: ordinary and structured chat, text/content
streaming, reasoning effort, model parameters, registered function tools,
async methods, callbacks, cancellation, cloning, rich image/PDF content,
history, echo, and response metadata. The compatibility implementation is
version-gated to exactly 0.4.2. The separate `parallel_chat*()` and
`batch_chat*()` helpers are outside the stable core contract. Parallel helpers
use the package blocker; batch helpers use ellmer's generic unsupported-provider
error because the Codex endpoint is stream-only. The underlying Codex transport
remains an undocumented external compatibility dependency.

The constructor and the returned Chat object are documented separately.
`?chat_codex` describes constructor arguments, credential side effects,
conditions, and model selection; the `ellmer` Chat help describes the methods
used for history, streaming, structured output, tools, callbacks, async work,
and cloning. The package does not claim compatibility with ellmer's separate
`parallel_chat*()` or `batch_chat*()` helpers because the subscription
transport is stream-only.

## Tool calling

Tools use the normal `ellmer` API. Codex can request one or more calls, the
package executes the registered functions locally, and the results are sent
back through the Responses protocol until Codex returns a final answer:

```{r, eval = FALSE}
weather_tool <- ellmer::tool(
  function(city) paste("Sunny in", city),
  name = "get_weather",
  description = "Get the current weather for a city.",
  arguments = list(city = ellmer::type_string())
)

chat <- chat_codex(model = Sys.getenv("ELLMERCODEX_MODEL"))
chat$register_tool(weather_tool)
chat$chat("What is the weather in Montevideo?")
```

Tool calls can be streamed as content, including their request and result
objects:

```{r, eval = FALSE}
coro::collect(chat$stream("Check the weather again.", stream = "content"))
```

Tool-containing turns stay in the history as `ellmer::ContentToolRequest` and
`ellmer::ContentToolResult` objects. `$chat_structured()` follows ellmer's
semantics and disables registered tools for that request; use `$chat()` first
when structured extraction needs tool-assisted context.

Structured output uses the normal ellmer type system:

```{r, eval = FALSE}
chat$chat_structured(
  "My name is Susan and I'm 13 years old.",
  type = ellmer::type_object(
    name = ellmer::type_string(),
    age = ellmer::type_integer()
  )
)
```

## Handle failures

The package uses structured condition classes so applications do not need to
parse messages. For example:

```{r, eval = FALSE}
tryCatch(
  chat$chat("Hello"),
  codex_auth_error = function(error) {
    message("Authentication failed; run codex_login() again.")
  },
  codex_chat_error = function(error) {
    message("The experimental Codex request failed: ", conditionMessage(error))
  },
  codex_protocol_changed_error = function(error) {
    message("The experimental upstream protocol may have changed.")
  }
)
```

Messages are sanitized before they are signaled. Raw response bodies, bearer
tokens, authorization codes, OAuth state, account identifiers, and
credential-store contents are not intentionally included.

For application code, handle the most specific condition first and then fall
back to the parent classes. `codex_auth_error` covers missing, malformed, and
expired credentials; `codex_request_error` covers transport failures; and
`codex_protocol_changed_error` indicates that the observed upstream response
shape may have changed. The full list is available in
`?ellmercodex-conditions`.

## Sign out

```{r, eval = FALSE}
codex_logout()
```

Logout removes only the package's own keyring entry and encrypted fallback
file. It does not remove credentials belonging to Codex CLI or other clients.
