---
title: "Getting started with ellmercodex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ellmercodex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`ellmercodex` connects `ellmer` chats to a ChatGPT subscription through
OpenAI's documented "Sign in with ChatGPT" flow for open-source, locally
hosted apps (<https://developers.openai.com/siwc/token-sharing-open-source>)
and the public Responses API. Its stable core is the public `ellmer` `Chat`
object (0.5.0 or later) for interactive, single-conversation work. ChatGPT
plan usage is an OpenAI preview feature; the package is not endorsed or
supported by OpenAI.

Package loading, examples, vignettes, and automated tests do not authenticate,
open a browser, inspect stored credentials, or make network requests.

## Check the local integration

```{r}
library(ellmercodex)

codex_available()
```

`codex_available()` checks installed dependencies, static configuration, and
the supported `ellmer` compatibility seam. Its default does not inspect a
credential store or contact the network, and `TRUE` does not guarantee that a
remote request will succeed. After authentication, `codex_models()` queries
`https://api.openai.com/v1/models` with the ChatGPT plan token and returns the
models listed for the account (`visibility: "list"`). Explicitly selected
model IDs absent from it are still passed to the service, which validates
access and reasoning effort:

```{r, eval = FALSE}
models <- codex_models()
models[c("id", "display_name", "default_reasoning_effort",
         "supported_reasoning_efforts")]
```

## Authenticate explicitly

Authentication happens only after a direct call to `codex_login()`. The
function opens a browser where you choose a ChatGPT account and workspace and
allow ChatGPT plan usage. It receives the callback on
`http://127.0.0.1:1455/callback` (set `ELLMERCODEX_CALLBACK_PORT` to change
the port), checks the OAuth state, exchanges the PKCE-protected code with the
client ID that OpenAI issued for this registration, verifies the ID token, and
checks that the `chatgpt.tokens.use.direct` scope was granted. It does not use
the macOS Keychain or another OS keyring.

```{r, eval = FALSE}
auth <- codex_login()
codex_account(auth)
```

`codex_account()` omits tokens and redacts the account identifier. The package
does not read the Codex CLI credential file or another application's stored
session.

Set `persist = FALSE` to keep the credential only in the current R process:

```{r, eval = FALSE}
auth <- codex_login(persist = FALSE)
chat <- chat_codex(model = Sys.getenv("ELLMERCODEX_MODEL"))
```

The process-local session is cleared by `codex_logout()` and its tokens are
never written to disk or an OS keyring.

Persistent credentials are stored in `credentials.json` below
`tools::R_user_dir("ellmercodex", "config")`, written atomically with
owner-only permissions. Set `ELLMERCODEX_HOME` before the first credential
access to use a different private directory:

```{r, eval = FALSE}
Sys.setenv(ELLMERCODEX_HOME = "/path/to/a/private/directory")
```

The same directory holds `host-id`, a stable and opaque host identifier that
OpenAI requires before the first sign-in. Access tokens last an hour and are
refreshed automatically; refreshes are serialized across R processes because
refresh tokens rotate. The first sign-in creates a registration bound to the
chosen account and workspace, and later sign-ins reuse it. To use another
account or workspace, call `codex_logout()` first. For an app hosted on a
Linux VM, follow the [single-process deployment guide](hosted-oauth.md).

## Create a text chat

By default, `chat_codex()` selects a usable model from the authenticated
account catalog. Select a model explicitly when you need a stable choice, and
choose reasoning effort from the catalog when it lists the model:

```{r, eval = FALSE}
chat <- chat_codex(
  system_prompt = "Be concise.",
  model = models$id[[1]],
  effort = models$supported_reasoning_efforts[[1]][[1]]
)

chat$chat("Remember that the codeword is amber. Reply: acknowledged")
chat$chat("What was the codeword?")
```

For example, if the account catalog omits the GPT-6 family but the chat service
accepts it, select a model explicitly:

```{r, eval = FALSE}
chat <- chat_codex(model = "gpt-6-luna", effort = "medium")
```

The returned object is an `ellmer` `Chat` (0.5.0 or later). Interactive Chat
methods remain available: ordinary and structured chat, text/content
streaming, reasoning effort, model parameters, registered function tools,
async methods, callbacks, cancellation, cloning, rich image/PDF content,
history, echo, and response metadata. The compatibility implementation checks
required ellmer contracts at runtime. Provider token counting and file
management are unsupported and fail before a request. The separate
`parallel_chat*()` and `batch_chat*()` helpers are outside the stable core
contract because ChatGPT plan usage is stream-only. ChatGPT plan usage also
rejects some Responses arguments, including `temperature`, `top_p`,
`max_tokens` (`max_output_tokens`), `metadata`, and `previous_response_id`;
the package fails before sending a request that uses them. Function tools are
sent in an `ellmer` tool namespace, as the preview requires.

The constructor and the returned Chat object are documented separately.
`?chat_codex` describes constructor arguments, credential side effects,
conditions, and model selection; the `ellmer` Chat help describes the methods
used for history, streaming, structured output, tools, callbacks, async work,
and cloning. The package does not claim compatibility with ellmer's separate
`parallel_chat*()` or `batch_chat*()` helpers because the subscription
transport is stream-only.

## Tool calling

Tools use the normal `ellmer` API. Codex can request one or more calls, the
package executes the registered functions locally, and the results are sent
back through the Responses protocol until Codex returns a final answer:

```{r, eval = FALSE}
weather_tool <- ellmer::tool(
  function(city) paste("Sunny in", city),
  name = "get_weather",
  description = "Get the current weather for a city.",
  arguments = list(city = ellmer::type_string())
)

chat <- chat_codex(model = Sys.getenv("ELLMERCODEX_MODEL"))
chat$register_tool(weather_tool)
chat$chat("What is the weather in Montevideo?")
```

Tool calls can be streamed as content, including their request and result
objects:

```{r, eval = FALSE}
coro::collect(chat$stream("Check the weather again.", stream = "content"))
```

Tool-containing turns stay in the history as `ellmer::ContentToolRequest` and
`ellmer::ContentToolResult` objects. `$chat_structured()` follows ellmer's
semantics and disables registered tools for that request; use `$chat()` first
when structured extraction needs tool-assisted context.

Structured output uses the normal ellmer type system:

```{r, eval = FALSE}
chat$chat_structured(
  "My name is Susan and I'm 13 years old.",
  type = ellmer::type_object(
    name = ellmer::type_string(),
    age = ellmer::type_integer()
  )
)
```

## Handle failures

The package uses structured condition classes so applications do not need to
parse messages. For example:

```{r, eval = FALSE}
tryCatch(
  chat$chat("Hello"),
  codex_auth_error = function(error) {
    message("Authentication failed; run codex_login() again.")
  },
  codex_chat_error = function(error) {
    message("The experimental Codex request failed: ", conditionMessage(error))
  },
  codex_protocol_changed_error = function(error) {
    message("The experimental upstream protocol may have changed.")
  }
)
```

Messages are sanitized before they are signaled. Raw response bodies, bearer
tokens, authorization codes, OAuth state, account identifiers, and
credential-store contents are not intentionally included.

For application code, handle the most specific condition first and then fall
back to the parent classes. `codex_auth_error` covers missing, malformed, and
expired credentials; `codex_request_error` covers transport failures; and
`codex_protocol_changed_error` indicates that the upstream response shape may
have changed. ChatGPT plan limits have their own subclasses:
`codex_usage_limit_error` (a `codex_rate_limit_error`) means the plan's usage
limit was reached, and `codex_usage_unavailable_error` means plan usage is
temporarily unavailable. The full list is available in
`?ellmercodex-conditions`.

## Sign out

```{r, eval = FALSE}
codex_logout()
```

Logout revokes the refresh token with OpenAI (pass `revoke = FALSE` to skip
this), then removes the package's own credential file and process-local
credential. It keeps the host identifier, and it does not remove credentials
belonging to Codex CLI or other clients.
