---
title: "Getting started with ellmercodex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ellmercodex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`ellmercodex` connects `ellmer` text chats to observed, subscription-backed
Codex authentication and response behavior. It is an independent experimental
integration. The direct OAuth and response transport are not a documented
third-party API and may change without notice.

Package loading, examples, vignettes, and automated tests do not authenticate,
open a browser, inspect stored credentials, or make network requests.

## Check the local integration

```{r}
library(ellmercodex)

codex_available()
models <- codex_models()
attr(models, "supported")
attr(models, "reason")
```

`codex_available()` checks installed dependencies, static configuration, and
the supported `ellmer` compatibility seam. Its default does not inspect a
credential store. Account-specific model discovery is not available through a
documented direct endpoint, so `codex_models()` returns a typed empty result
instead of an inferred catalog.

## Authenticate explicitly

Authentication happens only after a direct call to `codex_login()`. The
function opens a browser, validates a loopback callback using PKCE and OAuth
state, and stores the resulting credential in a package-specific OS-keyring
entry by default.

```{r, eval = FALSE}
auth <- codex_login()
codex_account(auth)
```

`codex_account()` omits tokens and redacts the account identifier. The package
does not read the Codex CLI credential file or another application's stored
session.

Set `persist = FALSE` to keep the credential only in the current R process:

```{r, eval = FALSE}
codex_login(persist = FALSE)
chat <- chat_codex(model = Sys.getenv("ELLMERCODEX_MODEL"))
```

The process-local session is cleared by `codex_logout()` and is never written
to disk or the OS keyring.

The OS keyring is the default backend. A file fallback is available only when
both of these environment variables are configured explicitly:

```{r, eval = FALSE}
Sys.setenv(
  ELLMERCODEX_CREDENTIAL_BACKEND = "file",
  ELLMERCODEX_CREDENTIAL_PASSPHRASE = "use-a-secret-from-your-password-manager"
)
```

The fallback is encrypted and stored below
`tools::R_user_dir("ellmercodex", "data")`. No plaintext fallback is provided.

## Create a text chat

Select a model explicitly because subscription eligibility is account- and
workspace-specific:

```{r, eval = FALSE}
chat <- chat_codex(
  system_prompt = "Be concise.",
  model = Sys.getenv("ELLMERCODEX_MODEL", "gpt-5.6-luna")
)

chat$chat("Remember that the codeword is amber. Reply: acknowledged")
chat$chat("What was the codeword?")
```

The returned object is an `ellmer` `Chat`. This release supports ordinary text
`$chat()`, text or content `$stream()`, and multi-turn history with `ellmer`
versions from 0.4.2 up to, but not including, 0.5.0. Structured output, tools,
asynchronous methods, and a native provider subclass are not part of the
supported surface.

## Handle failures

The package uses structured condition classes so applications do not need to
parse messages. For example:

```{r, eval = FALSE}
tryCatch(
  chat$chat("Hello"),
  codex_auth_error = function(error) {
    message("Authentication failed; run codex_login() again.")
  },
  codex_chat_error = function(error) {
    message("The experimental Codex request failed: ", conditionMessage(error))
  },
  codex_protocol_changed_error = function(error) {
    message("The experimental upstream protocol may have changed.")
  }
)
```

Messages are sanitized before they are signaled. Raw response bodies, bearer
tokens, authorization codes, OAuth state, account identifiers, and
credential-store contents are not intentionally included.

## Sign out

```{r, eval = FALSE}
codex_logout()
```

Logout removes only the package's own keyring entry and encrypted fallback
file. It does not remove credentials belonging to Codex CLI or other clients.
