---
title: "Getting started with ellmercodex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ellmercodex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`ellmercodex` connects `ellmer` text chats to observed, subscription-backed
Codex authentication and response behavior. It is an independent experimental
integration. The direct OAuth and response transport are not a documented
third-party API and may change without notice.

Package loading, examples, vignettes, and automated tests do not authenticate,
open a browser, inspect stored credentials, or make network requests.

## Check the local integration

```{r}
library(ellmercodex)

codex_available()
```

`codex_available()` checks installed dependencies, static configuration, and
the supported `ellmer` compatibility seam. Its default does not inspect a
credential store. After authentication, `codex_models()` queries the observed
account-specific catalog and returns the models and reasoning efforts the
account currently advertises:

```{r, eval = FALSE}
models <- codex_models()
models[c("id", "display_name", "default_reasoning_effort",
         "supported_reasoning_efforts")]
```

## Authenticate explicitly

Authentication happens only after a direct call to `codex_login()`. The
function opens a browser, validates a loopback callback using PKCE and OAuth
state, and stores the resulting credential in a package-specific OS-keyring
entry by default.

```{r, eval = FALSE}
auth <- codex_login()
codex_account(auth)
```

`codex_account()` omits tokens and redacts the account identifier. The package
does not read the Codex CLI credential file or another application's stored
session.

Set `persist = FALSE` to keep the credential only in the current R process:

```{r, eval = FALSE}
codex_login(persist = FALSE)
chat <- chat_codex(model = Sys.getenv("ELLMERCODEX_MODEL"))
```

The process-local session is cleared by `codex_logout()` and is never written
to disk or the OS keyring.

The OS keyring is the default backend. A file fallback is available only when
both of these environment variables are configured explicitly:

```{r, eval = FALSE}
Sys.setenv(
  ELLMERCODEX_CREDENTIAL_BACKEND = "file",
  ELLMERCODEX_CREDENTIAL_PASSPHRASE = "use-a-secret-from-your-password-manager"
)
```

The fallback is encrypted and stored below
`tools::R_user_dir("ellmercodex", "data")`. No plaintext fallback is provided.

## Create a text chat

Select a model explicitly because subscription eligibility is account- and
workspace-specific:

```{r, eval = FALSE}
chat <- chat_codex(
  system_prompt = "Be concise.",
  model = Sys.getenv("ELLMERCODEX_MODEL", "gpt-5.6-luna"),
  effort = "medium"
)

chat$chat("Remember that the codeword is amber. Reply: acknowledged")
chat$chat("What was the codeword?")
```

The returned object is an `ellmer` `Chat`. This release supports ordinary text
`$chat()`, text or content `$stream()`, ellmer `chat_structured()`, reasoning
effort, model parameters, registered function tools, and multi-turn history
with `ellmer` versions from 0.4.2 up to, but not including, 0.5.0.
Asynchronous methods and a native provider subclass are not part of the
supported surface.

## Tool calling

Tools use the normal `ellmer` API. Codex can request one or more calls, the
package executes the registered functions locally, and the results are sent
back through the Responses protocol until Codex returns a final answer:

```{r, eval = FALSE}
weather_tool <- ellmer::tool(
  function(city) paste("Sunny in", city),
  name = "get_weather",
  description = "Get the current weather for a city.",
  arguments = list(city = ellmer::type_string())
)

chat <- chat_codex(model = Sys.getenv("ELLMERCODEX_MODEL"))
chat$register_tool(weather_tool)
chat$chat("What is the weather in Montevideo?")
```

Tool calls can be streamed as content, including their request and result
objects:

```{r, eval = FALSE}
coro::collect(chat$stream("Check the weather again.", stream = "content"))
```

Tool-containing turns stay in the history as `ellmer::ContentToolRequest` and
`ellmer::ContentToolResult` objects. `$chat_structured()` follows ellmer's
semantics and disables registered tools for that request; use `$chat()` first
when structured extraction needs tool-assisted context.

Structured output uses the normal ellmer type system:

```{r, eval = FALSE}
chat$chat_structured(
  "My name is Susan and I'm 13 years old.",
  type = ellmer::type_object(
    name = ellmer::type_string(),
    age = ellmer::type_integer()
  )
)
```

## Handle failures

The package uses structured condition classes so applications do not need to
parse messages. For example:

```{r, eval = FALSE}
tryCatch(
  chat$chat("Hello"),
  codex_auth_error = function(error) {
    message("Authentication failed; run codex_login() again.")
  },
  codex_chat_error = function(error) {
    message("The experimental Codex request failed: ", conditionMessage(error))
  },
  codex_protocol_changed_error = function(error) {
    message("The experimental upstream protocol may have changed.")
  }
)
```

Messages are sanitized before they are signaled. Raw response bodies, bearer
tokens, authorization codes, OAuth state, account identifiers, and
credential-store contents are not intentionally included.

## Sign out

```{r, eval = FALSE}
codex_logout()
```

Logout removes only the package's own keyring entry and encrypted fallback
file. It does not remove credentials belonging to Codex CLI or other clients.
